Determinism, Complexity, and Predictability in Computer Performance

### 2014-08-01 response.tex

\emph{Page 19: ``There are some exceptions: [...] Conversely, forecast errors
for the \texttt{col\_major} and \texttt{dgesdd$_5$} signals are higher than the
corresponding WPEs suggest --- except for the nonlinear LMA predictor.'' This is
not really an exception. As mentioned by the authors before, in general, the
result only holds under the assumption that a reasonably good model has been
used for the prediction. But for ill-matched datasets and models like in the
cases mentioned above this is just not the case so it is not a surprise that the
forecast errors are higher.}

You are absolutely correct and we've adjusted the text accordingly.

\section*{Response to the Second Referee}

We thank the reviewer for taking the time to read our paper and
provide us with valuable feedback.  The questions raised in this
review caused us to completely rethink the results, the fits, the
findings, and the conclusions---hence the long delay in resubmitting
this new version, and the length and depth of this response letter.

@@

\noindent The dashed curve in this plot (and in Figure 6) was fit
\emph{solely on the computer performance data}.  Nonetheless, the
points from the other time series fall within a narrow band around the
fit.  Note that this fit ($y = a \log(b x + 1)$) is different than in
the submitted version.  A careful examination of the data and some
careful thought about the meaning and theoretical bounds of
MASE---prompted by the reviewer's questions---suggested that this was
a better match to the data.  This analysis and consideration is
chronicled in the third paragraph of the results section.  Finally,
note that there are a lot more points on this plot than in our
original version; this is discussed further below.

\smallskip

\emph{1) Introduction is too long, but the main problem is not
  formulated definitely enough. I see that the main point is the
  approximately linear (or at least one-to-one) relationship between
  WPE and the best prediction error. This aim is not claimed
  explicitly in the Introduction. All other aspects of the manuscript
  do not seem important enough to be published in a separate article.}

We have shortened the introduction and modified the manuscript
throughout to make it more clear that our goal was to derive a
practical criterion for best-case prediction error in a model-free
way, and to test it in the context of empirical data.  We have also
changed the title to line up with this goal.

@@

\smallskip

\emph{2) In particular, two primary findings mentioned on page 2
  (Introduction) are formulated quite vaguely. The first part of the
  first finding reads ``(i) complexity of a noisy real-value time
  series is quantifiable by permutation entropy''. However, it seems
  obvious from the previous research, including that cited by the
  authors. The authors do not suggest PE or WPE here, they exploit
  them as previously known approaches. It was previously known that PE
  is an estimate of KS entropy as the authors also state. It was well
  known that KS entropy relates to predictability and, in this sense,
  it is a measure of complexity (see e.g. the work [G. Boffetta,
    M. Cencini, M. Falcioni, and A.  Vulpiani, ``Predictability: a way
    to characterize complexity'' // Physics Reports 356 (2002)
    367–474] which is not cited by the authors). Thus, the first part
  of the first finding does not provide any new information. The
  second part of the first finding reads ``(ii) complexity of the
  noisy real-valued time series is correlated with prediction accuracy
  of an appropriate predictor''.  However, it seems also obvious and
  directly follows from the definition of KS- entropy and its
  discussion in many papers including [Boffetta et al.] cited above.}

These claims have indeed been explored extensively in theory and in
toy models.  Our specific goal, which was not well explained in the
submitted version, was to show that those claims also hold---and can
be practically applied---in real-world time-series data that may
contain dynamical and/or observational noise.  This has not, to our
knowledge, received much attention in the literature.  We have added a
paragraph to the introduction that hopefully makes all of this more
clear.

We appreciate being informed about the Boffetta et al.  paper, and we
have included a discussion of it at the end of Section II.

@@

\smallskip

\emph{3) By the way, the authors call ``fully complex'' the white
  noise process, i.e. they equate predictability and
  complexity. However, they do not even mention another approach to
  the notion of complexity which ascribes low complexity to the white
  noise [the works e.g. by C.R. Shalizi and J. Crutchfield carried out
    in the Santa Fe Institute from where one of the authors is]. It
  would be appropriate to mention that ambiguity of the notion of
  complexity if the latter plays an important role in many
  formulations of the authors.}

We have added a short discussion about the differences in notions and
definitions of complexity to Sections I and II and clarified the
working definition that we use in our work.

@@

\smallskip

\emph{4) The second finding reads ``The way information is generated
  and processed internally by a system plays a crucial role in the
  success of different forecasting schema - and in the choice of which
  one is appropriate for a given time series''. This statement also
  looks trivial. Indeed, an optimal predictor for a linear stochastic
  system and low-dimensional nonlinear deterministic system are quite
  different (an AR model versus a local or global nonlinear model) due
  to different properties of the original systems. Thus, there is
  nothing to be proven here. Probably, the authors imply their
  particular benchmark signals (reflecting a computer behavior) but
  they do not claim that.  Thus, both primary findings seem to provide
  nothing new.}

The reviewer is absolutely right.  This finding, as it was originally
stated, is indeed trivial.  Our intent, as we have modified the text
to reflect, is that the appropriateness of a predictor for a given
time series can be discovered \emph{empirically}, via comparison of
its MASE value with the WPE value of the data.  Previous explorations
of predictability have used clean data and/or systems where the
generating process is known.  Our work makes neither of these
assumptions.

@@

Given a linear stochastic system, a linear stochastic prediction model
should certainly be more productive than a nonlinear deterministic
prediction model.  But in real-world data, the generating process is
almost never known, and it can be very hard to evaluate a model that
one has constructed.  Standard methods like the AIC approach used in
ARIMA modeling make a choice from among a large class of possible
models.  These selection algorithms rarely have a ``none of these are
appropriate'' output; they simply choose the model that best satisfies
their built-in criteria---even if none of the models in the class are
really appropriate.

The challenge that this poses for a practitioner is to know whether
there is a mismatch between a particular time series from an unknown
system and a particular forecast model: i.e., whether one can do
better and should try another model, or whether the time series is so
complicated that the model under consideration is doing as well as one
could expect.  The curves and shaded regions on the WPE vs. MASE plots
in our paper are an empirical heuristic that should help practitioners
do that.

@@

\smallskip

\emph{5) The benchmark system - a computer as an object of prediction
  - seems to be an unexpected and, probably, interesting choice but
  just as an additional illustration of a statement which should be
  first shown for well-controlled and well-understood paradigmatic
  systems (like low-dimensional maps or ordinary differential
  equations). Indeed, it is not known in advance what error one should
  expect for such a complicated object using a certain predictive
  technique, what technique is optimal or close to optimal, etc. If
  the main point of the research is the relationship between WPE and
  the best prediction error (if any), selection of such a complicated
  object as computer performance variations only introduces its own
  difficulties into the problem.}

Computers are indeed complicated objects, but we have studied their
dynamics extensively and established quite firmly that they are
garden-variety dynamical systems with surprisingly low-dimensional
dynamics [Mytkowicz et al. CHAOS, 2009].  Standard nonlinear
time-series analysis techniques on performance data measured from
running computers produce clean and consistent results, as mentioned
above: fixed points, bifurcations, chaotic attractors, and so on.

Well-known systems are certainly good test cases, but complicated,
unknown systems are where prediction really matters---especially
systems where a good prediction can make a practical difference (e.g.,
reallocation of resources in a computer to suit the dynamically
changing needs of a program).  The relationship posited in our paper
does indeed hold for paradigmatic systems like the H\'{e}non map and
the Lorenz equations.  We had not included these examples in our paper
because they simply confirmed well-known results, and because our
focus was on confirming those results in the context of noisy,
real-world data.  The penultimate paragraph of the results section now
explains that the WPE vs. MASE relationship for these two synthetic
examples (as well as for the Santa Fe ``A'' data set, for a
white-noise process, and for various nonlinear transformations of our
original data) all follow the empirically determined pattern in our
paper.

@@

\smallskip

\emph{6) The description of the choice of the benchmark system in the
  Introduction is too long. On the other hand, if the computer signals
  are the main interest of the work, then it should be claimed
  explicitly. Then, the paper should be rewritten and submitted to a
  kind of engineering journals where it might well be of interest.}

The primary goal of this paper was to explore the relationship between
prediction error and WPE in noisy, real-world systems, and to assess
the practical applicability of this relationship.  Accordingly, we
have cut down on the level of detail regarding the benchmark system
and pointed interested readers to our other papers for the engineering
details.  We did not feel that removing all of this material would be
effective, however, because it helps the reader understand why this is
a good testbed for exploring the problem at hand.

@@

\smallskip

\emph{7) As for the choice of the four prediction techniques, some of
  them seem superfluous. E.g. naive approach is just a global AR model
  of the zero order.  Random-walk is similar to the AR model of the
  first order where the coefficient is not estimated by the
  least-squares but set equal to 1 (or this is an ARIMA model with the
  first difference and ``zero ARMA part''). Inefficiencies of the
  ARIMA model as compared to those choices in the paper can represent
  an inappropriate procedure for the order selection rather than
  indicate inappropriateness of the entire ARIMA approach. Thus,
  linear ARIMA and nonlinear LMA might be sufficient (in combination
  with mathematical benchmark examples) to represent close-to-optimal
  prediction errors.}

It is certainly true that the naive and random-walk predictors can be
seen as subclasses of the ARIMA model family.  We have explicitly
included them, however, because they are used (sometimes to great
success) in practice, because that made for a more-comprehensive
exploration, and because of a subtle issue regarding model building
procedures.

@@

This short paragraph in this paper, which explains the efficiencies
and inefficiencies of these schemes, was intended to serve two
purposes: first, to explain why we have included these simple
prediction schemes and not just the more-complicated ones; second, to
help in explaining the specific results described later in the paper.

\smallskip

\emph{9) A better justification of the prediction error metrics could
  be also in order if this work is continued. At least, comparison of
  prediction errors quantified with different metrics seems
  necessary. Especially, taking into account the following major
  concerns.}

We have extended the justification for our use of the MASE metric; see
the last paragraph in the results section.  As mentioned there, the
paper in which MASE was introduced---our reference [21]---includes an
extensive evaluation of this metric, as well as a comprehensive
comparison to other metrics.  If the reviewer thinks that it would
strengthen our paper, we can recapitulate more of that material in our
paper.  As part of our current work on this topic, we are evaluating
how other choices of metric affect the shape of our curves, but we
feel that a comprehensive study of this issue is best left as a sequel
to this work.

@@

\smallskip

\noindent\emph{General remarks.}

\emph{I) Strict relationship between WPE and prediction error
  applicable to all systems is hardly possible. In particular, there
  is a notion of epsilon-entropy which tends to the KS entropy if
  epsilon (the cell size) tends to zero (see e.g.  [Boffetta et al.]
  cited above). For stochastic systems, the KS entropy is infinite and
  epsilon-entropy varies strongly with the scale epsilon. Seemingly,
  the permutation entropy estimate in this case will strongly depend
  on the choice of the word length L. The authors concentrate on L = 6
  and obtain coefficients of the regression between WPE and prediction
  error only for that choice and for their particular
  signals. However, other systems and other choices of L may easily
  lead to other regression coefficients so that deciding about the
  best possible prediction error from WPE seems impossible in
  general. It is good if such a decision is possible for a certain
  narrow class of systems. However, the authors do not even discuss to
  which class of systems their results apply.}

WPE values depend not only on wordlength, but also on \emph{data}
length, since one needs more data to get good statistics on the words
for the PE calculation if those words are longer.  These two
interacting knobs are not easy to set.  Perhaps as a result of this,
wordlength choices are rarely justified in the permutation entropy
literature.  The extent of the discussion in [Bandt \& Pompe, {\sl
    PRL}, 2002], for instance---the paper in which PE was
proposed---is ``Our entropies are calculated for different [$\ell$]
but we do not determine a limit for large [$\ell$], although this is
an interesting theoretical problem. For practical purposes, we
recommend 3\dots7...''  The corresponding discussion in [Cao {\sl et
    al.}, {\sl PRE}, 2004] reads ``In their paper, Bandt \& Pompe
recommended [$\ell$] of 3-7.  We often found that 3 and 4 may still be
too small, and a value of 5, 6, or 7 seems to be the most
suitable....''

The right thing to do, as we argued in the last paragraph of Section
V, is to compute PE across a \emph{range} of wordlengths ($\ell$) and
look for a flat spot in the curve of PE versus $\ell$.  We applied
that approach individually to each of the time series studied here to
choose a $\ell$ value that was appropriate to that particular trace.
(All of those values ended up being 6, which is consistent with the
informal arguments in previous papers, but that was not an \emph{a
  priori} or global choice.)  We conjecture that a systematic approach
like this will mitigate any variation in the results, thereby
preserving the shape of the WPE vs. MASE curve.  We have not yet
confirmed that conjecture, but we are in the process of doing so.

Regarding generality: the goal of this paper was to explore model-free
heuristics for empirical time-series data, where one does not know
anything about the underlying system.  In situations like this,
equivalence classes are difficult to define.  We addressed
this---insofar as is possible in an experimental context---by choosing
a range of time-series data sets whose behavior spans a wide range of
dynamical behaviors.  For all of these examples---as well as for the
49 additional ones
% \footnote{H\'{e}non, Lorenz, SFI A, 
%  nonlinear transformations of the 45 svdone, svdtwo and signals}
constructed as a result of the reviewer's questions and described in
different parts of this response letter---the pattern in Figure 7
remained virtually unchanged.  This strongly suggests, but of course
does not prove, that our approach is general.  As the reviewer
suggested, this will require extensive study and evaluation in future
work.

@@

Regarding the first part of this point: while PE is invariant under an
invertible nonlinear transformation of the values in the time series,
WPE is not---which amplifies the potential concerns about this issue.
To explore this, we performed the invertible nonlinear transformation
$f(x)=5e^{x(t)}$ on several of our traces ($x(t)=$ {\tt dgesdd}$_2$,
{\tt dgesdd}$_5$, and {\tt dgesdd}$_6$).  The image below shows WPE
vs. MASE results of the best predictor on these three traces, along
with those on the untransformed originals:

\begin{center}
    \includegraphics[width=0.8\columnwidth]{figs/nonlinear_transform_data}
\end{center}

The WPE vs. MASE points of the transformed data move---upwards,
generally---but they remain within the shaded region.  Indeed, the
values for the transformed version of {\tt dgesdd}$_5$ (dark-green
icons) move {\sl closer} to the curve.  When these 45 new points are
included in the $y = a \log(b x + 1)$ fit, the coefficients
\alert{barely change: $a$ by n\% and $b$ by n\%.  Ryan, what are these
  numbers?}

This series of experiments supports our claims of generality: a
transformation like this effectively changes the generating process,
but does not affect the performance of our model-free heuristic.  And
the curves that capture that heuristic---which were derived from data
from a single computer-performance experiment---do not only hold for
traces from that experiment and transformed versions of them.  The
preliminary results of our broader study, currently underway, indicate
that they also hold for data from synthetic examples (the H\'{e}non
map, the Lorenz equations, and the output of a purely stochastic
random-walk process) and for real-world data from a different system
(SFI data set A).  The image below shows some of these results:

@@

\begin{center}
    \includegraphics[width=0.8\columnwidth]{figs/new_prediction_vs_entropy_extras_with_nonlinear_all_points}
\end{center}

The blue fit lines here are derived from the original 120 traces; the
magenta one is fit to those traces plus the nonlinear transformations
of {\tt dgesdd}$_2$, {\tt dgesdd}$_5$, and {\tt dgesdd}$_6$, the $x$
coordinate of a 10,000-point orbit from the H\'{e}non map with $a=1.4$
and $b=0.3$, the $x$ coordinate of a 10,000-step 4th-order Runge-Kutta
solution of the Lorenz equations with $r=28, \; \sigma=10, \; b=8/3$
and $\Delta t=0.01$, a 10,000-point random-walk time series generated
\alert{how, Ryan?}, and data set ``A'' from the Santa Fe Institute
time-series competition website.

Note that the inclusion of these additional 49 points hardly changed
the main central curve of the fit.  Indeed, doing so caused the
one-$\sigma$ volume of that fit to shrink (viz., the dashed magenta
curves inside the dashed blue curves).  That is, the additional points
firmed up the findings.  The magenta curve reflects a map, a flow, a
purely stochastic signal, a variety of real-world data from two
different systems (laser, computer), and nonlinear transformations of
that real-world data---again, supporting the generality of our
heuristic.

We are in the process of broadening this study to other systems,
transformations, and situations; in the meantime, we have added a
sentence about these preliminary results to the penultimate paragraph
in Section VI.

### 2014-08-01 results.tex
synching everything

The discussion in the previous paragraph highlights the second finding
of this paper: the ability of our proposed heuristic to flag
inappropriate models.  \arima is an automated, mechanical procedure
for choosing parameters for an ARIMA model of a given data set.  While
the tests and criteria employed by this algorithm are sophisticated
(see Section~\ref{sec:arima}), the results can still be
sub-optimal---if the initial space of models being searched is not
broad enough, for instance, or if one of the preliminary tests gives
an erroneous result.  \arima \emph{always} returns a model, and it can
be very hard to detect when that model is bad.  Our results suggest a
way to do so: if the MASE score of an auto-fitted model like \arima is
out of line with the WPE value of the data, that can be an indication
of inappropriateness in the order selection and parameter estimation
procedure.  A close examination of the \arima MASE scores and WPE
values of the 15 \col traces, for instance, brought out such a
situation: all of these signals had the same complexity, but some (the
outliers in the blue cluster of $+$ points in the bottom-left panel of
Figure~\ref{fig:what}) had much higher \arima MASE values, indicating
a failure in the \arima process.  We then went back and manually
assisted the automated fitting procedure---forcing inclusion of a term
that is appropriate for a nonstationary signal.  In all cases, this
produced ARIMA models with MASE scores in line with the others in that
cluster.

@@

The second effect that can skew MASE scores is nonstationarity.  Since
this metric is normalized by the error of a random-walk forecast
\emph{on the training signal}, differences between the test signal and
training signal can create issues.  This is why the MASE values in
Table~\ref{tab:error} are not identically one for every random-walk
forecast of every time series: the last 10\% of these signals is
significantly different from the first 90\%.  The deviation from 1.00
will depend on the process---whether it has multiple regimes, what
those regimes look like, and how it switches between them---as well as
the experimental setup (e.g., sensor precision and data length).  For
the processes studied here, these effects do not cause the MASE values
to exceed 1.15, but pathological situations (e.g., a huge switch in
scale right at the training/test signal boundary) could produce higher
values.  This suggests another potentially useful heuristic: if the
MASE of a random-walk prediction of a time series is significantly
different from 1, this could be an indication that the signal is
nonstationary.  We are in the process of exploring this idea.

>>>>

The discussion in the previous paragraph highlights the second finding
of this paper: the ability of our proposed heuristic to flag
inappropriate models.  \arima is an automated, mechanical procedure
for choosing parameters for an ARIMA model of a given data set.  While
the tests and criteria employed by this algorithm
(Section~\ref{sec:arima}) are sophisticated , the results can still be
sub-optimal---if the initial space of models being searched is not
broad enough, for instance, or if one of the preliminary tests gives
an erroneous result.  \arima \emph{always} returns a model, and it can
be very hard to detect when that model is bad.  Our results suggest a
way to do so: if the MASE score of an auto-fitted model like \arima is
out of line with the WPE value of the data, that can be an indication
of inappropriateness in the order selection and parameter estimation
procedure.  A close examination of the \arima MASE scores and WPE
values of the 15 \col traces, for instance, brought out such a
situation: all of these signals had the same complexity, but some (the
outliers in the blue cluster of {\color{blue}$+$} points in the
bottom-left panel of Figure~\ref{fig:wpe_vs_mase_all}) had much higher
MASE values, indicating a failure in the \arima process.  We then went
back to the data, noticed some shifts of the baseline in those
signals, and manually assisted the automated fitting procedure to
force the inclusion of a term that is appropriate for a nonstationary
signal.  In all cases, this produced ARIMA models with MASE scores in
line with the others in that cluster.

@@

The second effect that can skew MASE scores is nonstationarity.  Since
this metric is normalized by the error of a random-walk forecast
\emph{on the training signal}, differences between the test signal and
training signal can create issues.  This is why the MASE values in
Table~\ref{tab:error} are not identically one for every random-walk
forecast of every time series: the last 10\% of these signals is
significantly different from the first 90\%.  The deviation from 1.00
will depend on the process---whether it has multiple regimes, what
those regimes look like, and how it switches between them---as well as
the experimental setup (e.g., sensor precision and data length).  For
the processes studied here, these effects do not cause the MASE values
to exceed 1.15, but pathological situations (e.g., a huge switch in
scale right at the training/test signal boundary, or a signal that
simply grows exponentially) could produce higher values.  This
suggests another potentially useful heuristic: if the MASE of a
random-walk prediction of a time series is significantly different
from 1, this could be an indication that the signal is nonstationary.
We are in the process of exploring this idea.
